---
title: "cebpd_crisprOCIAML5_DESeq_analysis"
author: "SP"
date: "3/9/2022"
output: html_document
---

##Load packages

```{r}
# R packages
library(digest)
library(rlang)
library(Rcpp)
library(htmltools)
library(glue)
library(ellipsis)
library(fansi)
library(vctrs)
library(tibble)
library(dplyr)
library(lattice)
library(BiocManager)
library(pkgload)
library(tidyverse)
library(renv)
library(here)

# Bionconductor
library(S4Vectors)
library(IRanges)
library(DelayedArray)
library(SummarizedExperiment)
library(DESeq2)
library(fgsea)
library(gage)
library(AnnotationHub)
library(EnhancedVolcano)

```

##Set gene counts dataset directroy path

```{r}
directory <- here::here() #here::here() represents the current working directory
getwd()
```

##Generate HTSeqsampleTable

```{r}
samplenames <- c("NS7_2", "NS8_1", "NS9_1", "NS10_1", "CEBPD7_1", "CEBPD8_1", "CEBPD9_1", "CEBPD10_2")
samplefiles <- c("NS7_2.gene.counts.txt", "NS8_1.gene.counts.txt", "NS9_1.gene.counts.txt", "NS10_1.gene.counts.txt", "CEBPD7_1.gene.counts.txt", "CEBPD8_1.gene.counts.txt", "CEBPD9_1.gene.counts.txt", "CEBPD10_2.gene.counts.txt")
sampleconditions <- c("NS", "NS", "NS", "NS", "CEBPD", "CEBPD", "CEBPD", "CEBPD")
sampleTable <- data.frame(sampleName = samplenames, fileName = samplefiles, condition = sampleconditions)
sampleTable$condition <- factor(sampleTable$condition)
```

##Generate DESeqDataSet

```{r}
ddsHTseq <- DESeqDataSetFromHTSeqCount(sampleTable = sampleTable, directory = directory, design = ~ condition)
```

##Combine gene names from gtfv39 to ddsHTseq

```{r}
#load gencode.key.Rdata
base::load(here::here("gencode.key.Rdata")) #somehow load() doesn't work
head(gencode.key)

#check gencode.key and ddsHTseq dimensions
dim(gencode.key)
dim(ddsHTseq)

#Do matchup ENSEMBL IDs?
all(rownames(ddsHTseq) == gencode.key$id)

#matchup ENSEMBL IDs
gencode.key <- gencode.key[match(rownames(ddsHTseq), gencode.key$id),]

#check if ENSEMBL IDs matchup now
all(rownames(ddsHTseq) == gencode.key$id)

#Pre-check rowData ddsHTseq for number of columns it has
rowData(ddsHTseq)

#add gene names from gencode.key to ddsHTseq 
mcols(ddsHTseq) <- cbind(mcols(ddsHTseq), gencode.key)

#Re-check rowData ddsHTseq for number of columns it has
rowData(ddsHTseq)
```

##Set control sample (NS) as first factor level

```{r}
ddsHTseq$condition <- factor(ddsHTseq$condition, levels = c("NS", "CEBPD"))
```

##View ddsHTSeq

```{r}
ddsHTseq
summary(ddsHTseq)
colData(ddsHTseq)
rowData(ddsHTseq)
assay(ddsHTseq)
dim(ddsHTseq)
```

##Filtering data for more than 4 reads

```{r}
keep4 <- rowSums(counts(ddsHTseq)) >= 4
ddsHTseq <- ddsHTseq[keep4, ]
table(keep4)
summary(ddsHTseq)
```

##DESeq

```{r}
dds <- DESeq(ddsHTseq)
resultsNames(dds)
```

##plotPCA

```{r}
vsd <- vst(dds, blind = FALSE)
plotPCA(vsd)
colnames(dds)

rld <- rlog(dds, blind = FALSE)
plotPCA(rld)

p = plotPCA(rld)

p = p + geom_label(label = names)
print(p)
```

##dds_results

```{r}
dds_res <- results(dds)
dim(dds_res)
```

##Add new ddsHTseq column name **gene** to dds_res

```{r}
#check ddsHTseq column names
mcols(ddsHTseq)

#add new ddsHTseq column **gene** to dds_res
dds_res$gene <- mcols(ddsHTseq)$gene

#check dds_res column names
colnames(dds_res)
```

##plotMA

```{r}
plotMA(dds_res, xlab = "Mean of normalized counts", ylab = "log2FoldChange", main = "CEBPD vs NS", alpha = 0.05)
```

Enhanced volcano for CEBPD manuscript

y = "padj" plots the -log10(padj) values on y-axis. That's why ylab = -log10(padj). -log10(0.05) equals to 1.3 fold change that's what we are seeing in the plot.

```{r}
#dds_res$gene <- row.names(dds_res)

EnhancedVolcano(dds_res, lab = dds_res$gene, x = "log2FoldChange", y = "padj", axisLabSize = 12, xlim = c(-9, 9), pCutoff = 0.05, FCcutoff = 0, labSize = 3, xlab = "log2FoldChange", ylab = "-log10(padj)", ylim = c(0, 8), legendPosition = "top", legendLabSize = 11, legendIconSize = 3.0,labFace = 'bold', gridlines.major = FALSE, gridlines.minor = FALSE)
```

##FGSEA

```{r}
#rank genes based on log2FC
gene.list = dds_res$log2FoldChange
names(gene.list) <- dds_res$gene
gene.list = sort(gene.list, decreasing = TRUE)

#removing NAs & duplicates
gene.list = gene.list[!is.na(gene.list)]
gene.list = gene.list[!duplicated(names(gene.list))]
head(gene.list)

msigdb.url = 'https://data.broadinstitute.org/gsea-msigdb/msigdb/release/7.5/'

system('wget https://data.broadinstitute.org/gsea-msigdb/msigdb/release/7.5/h.all.v7.5.symbols.gmt')

GO_file = 'h.all.v7.5.symbols.gmt'
myGO = gmtPathways(GO_file)

fgRes = as.data.frame(fgsea(pathways = myGO, stats = gene.list, minSize=5, maxSize=600, nperm=10000))

fgRespvalorder <- fgRes[fgRes$pval <= 0.05,][order(fgRes$NES[fgRes$pval <= 0.05], decreasing =TRUE),]
fgRespvalorder

fgRespvalorder <- as.data.frame(fgRespvalorder)
fgRespvalorder <- fgRespvalorder[, -8]

write.csv(fgRespvalorder, file = "fgsea_log2FConly_results_44.csv")
```

FGSEA barplot: fgsea_log2FConly_results_44_mod.csv

```{r}
gseaplot <- read.csv("fgsea_log2FConly_results_44_mod.csv")
gseaplot0.05 <- filter(gseaplot, padj < 0.05)

ggplot(gseaplot0.05, aes(x = NES, y = reorder(pathway, NES))) + geom_col(fill = "skyblue") + theme_classic()
```
