---
title: "Untitled"
author: "SP"
date: "1/27/2023"
output: html_document
---

```{r}
library(tidyverse)
library(methylKit)
library(pheatmap)
library(RColorBrewer)
library(ggpubr)
```

Filter and CEBPD log2fc data

```{r}
rvdlog2fc <- read.table("DESeq2_Result_RvsD_logFC_and_qvalue.txt", header = T)
str(rvdlog2fc)
unique(rvdlog2fc$patientID)

rvdCEBPDlog2fc <- rvdlog2fc %>% filter (geneID == "CEBPD") %>% dplyr::select(patientID, RvsD_logFC)
col_names <- rvdCEBPDlog2fc$patientID
rvdCEBPDlog2fc <- t(rvdCEBPDlog2fc)
colnames(rvdCEBPDlog2fc) <- col_names

rvdCEBPDlog2fc <- as.data.frame(rvdCEBPDlog2fc)
rvdCEBPDlog2fc <- rvdCEBPDlog2fc[2,]
row.names(rvdCEBPDlog2fc) <- "CEBPD"

rvdCEBPDlog2fc
#write.csv(rvdCEBPDlog2fc, file = "rvdCEBPDlog2fc.csv")
```

Renaming Rdata files coherent to expression data samples.....LATER EXECUTION NOT NEEDED

```{r}
#cpg.PAML_UHU samples
lst <- list.files(pattern = "cpg.PAML_UHU")
for (i in lst) {
  print(paste0("Renaming", i))
  file.rename(i, paste0(strsplit(i, "[._]")[[1]][2], strsplit(i, "[._]")[[1]][4], ".Rdata"))
}

#cpg.Sample_PAML_UHU samples
lst <- list.files(pattern = "cpg.Sample_PAML_UHU")
for (i in lst) {
  print(paste0("Renaming", i))
  file.rename(i, paste0(strsplit(i, "[._]")[[1]][3], strsplit(i, "[._]")[[1]][5], ".Rdata"))
}

#methylCall_Samples
lst <- list.files(pattern = "methylCall_Sample")
for (i in lst) {
  print(paste0("Renaming", i))
  file.rename(i, paste0(strsplit(i, "[._]")[[1]][3], strsplit(i, "[._]")[[1]][4], strsplit(i, "[._]")[[1]][5], ".Rdata"))
}

#cpg.PAML samples
lst <- list.files(pattern = "cpg.PAML")
for (i in lst) {
  print(paste0("Renaming", i))
  file.rename(i, paste0(strsplit(i, "[._]")[[1]][2], strsplit(i, "[._]")[[1]][3], ".Rdata"))
}
```

Keep only renamed Rdata files matching expression sample IDs.....LATER EXECUTION NOT NEEDED

```{r}
#Expression samples list
exp.names <- names(rvdCEBPDlog2fc)
exp.rda <- character()

for (i in 1:length(exp.names)) {
  p <- exp.names[i]
  exp.rda[i] <- paste0(p, ".Rdata")
}
exp.rda

#Identify rda file for which there is no expression data
methrda <- list.files(pattern = ".Rdata")

methrda_diff <- setdiff(methrda, exp.rda) #names present in vector1, but not in vector2
methrda_diff
length(methrda_diff)

#Delete file without expression data from the working directory
unlink(methrda_diff)

#Check if all the methrda files match with exp.rda names
methrda <- list.files(pattern = ".Rdata")
setdiff(methrda, exp.rda)
```

Filter CEBPD gene's promoter all DMCs.....LATER EXECUTION NOT NEEDED

```{r}
filenames <- list.files(full.names = F, pattern = "*.Rdata")
outputnames <- str_remove(filenames, ".Rdata")

for (i in 1:length(filenames)) {
  load(filenames[i])
  d <- as(myDiff10p, "data.frame")
  dmc <- d %>% 
    filter(chr == "chr8") %>%
    filter(start > 48650520) %>% 
    filter(end < 48651270) 
  dmc <- dmc %>% rename_at("meth.diff", ~outputnames[i])
  write.csv(dmc, paste0(outputnames[i], ".csv"))
}
```

Merging all patients CEBPD DMCs data.....LATER EXECUTION NOT NEEDED

```{r}
pAML_files <- list.files(full.names = F, pattern = "*.csv" )
output_files <- str_remove(pAML_files, ".csv")

c <- read.csv("PAML1.csv")
c <- c %>% 
  dplyr::select(contains("PAML") | contains("SGU"), "start")
for ( i in 2:length(pAML_files)){
  temp_data <- read.csv(pAML_files[i])
  temp_data <- temp_data %>% 
    dplyr::select(start, contains("PAML") | contains("SGU"))
  c <- merge(c, temp_data, by = "start", all.x = T, all.y = T)  #check all.x and all.y
  write.csv (c, paste0("pAML_DMCs.csv"))
}
```

Dendogram

```{r}
d <- read.csv("pAML_DMCs.csv")
d <- d[colSums(!is.na(d)) > 0]
d[is.na(d)] <- 0
dim(d)

f <- d
rownames(f) <- f$start

f <- f %>% dplyr::select(-X)
f <- f %>% dplyr::select(-start)
g <- as.matrix(f)
class(g)

#hclust
plot(hclust(dist(f)))
plot(hclust(dist(g)))
```

Annotating heatmap

```{r}
##Without sample PAML59 as it appears to be an outlier
rvdCEBPDlog2fc
rvdCEBPDlog2fc_t <- t(rvdCEBPDlog2fc)
rvdCEBPDlog2fc <- as.data.frame(rvdCEBPDlog2fc_t)
rvdCEBPDlog2fc$CEBPD <- as.numeric(rvdCEBPDlog2fc$CEBPD)

rvdCEBPDlog2fc[rvdCEBPDlog2fc$CEBPD < 0,] <- -2
rvdCEBPDlog2fc[rvdCEBPDlog2fc$CEBPD > 0,] <- 2
rvdCEBPDlog2fc <- subset(rvdCEBPDlog2fc, !row.names(rvdCEBPDlog2fc) %in% c("PAML59"))

g <- g[, colnames(g) != "PAML59"]
colnames_g <- colnames(g)
rvdCEBPDlog2fc <- subset(rvdCEBPDlog2fc, row.names(rvdCEBPDlog2fc) %in% colnames_g)

anno_color <- list(group = brewer.pal(3, "Set2"))
anno_color <- list(CEBPD = c("-2" = "cornflowerblue", "2" = "coral1", "0" = "bisque2"))
pheatmap(g, annotation = rvdCEBPDlog2fc, annotation_colors = anno_color, fontsize = 15)

```

CEBPD expression vs DMC correlation

```{r}
rvdlog2fc <- read.table("DESeq2_Result_RvsD_logFC_and_qvalue.txt", header = T)
class(rvdlog2fc)
names(rvdlog2fc)
unique(rvdlog2fc$patientID)

rvdCEBPDlog2fc <- rvdlog2fc %>% filter (geneID == "CEBPD") %>% dplyr::select(patientID, RvsD_logFC)
col_names <- rvdCEBPDlog2fc$patientID
rvdCEBPDlog2fc <- t(rvdCEBPDlog2fc)
colnames(rvdCEBPDlog2fc) <- col_names
class(rvdCEBPDlog2fc)

rvdCEBPDlog2fc <- as.data.frame(rvdCEBPDlog2fc)
rvdCEBPDlog2fc <- rvdCEBPDlog2fc[2,]
row.names(rvdCEBPDlog2fc) <- "CEBPD"


colmeans_g <- as.data.frame(colMeans(g))
colnames(colmeans_g) <- "DMC"

rvdCEBPDlog2fc <- as.data.frame(t(rvdCEBPDlog2fc))

names_g <- row.names(colmeans_g)
rvdCEBPDlog2fc <- subset(rvdCEBPDlog2fc, row.names(rvdCEBPDlog2fc) %in% names_g)

dim(colmeans_g)
dim(rvdCEBPDlog2fc) 

cebpd_exp_dmc_data <- cbind(colmeans_g, rvdCEBPDlog2fc)
str(cebpd_exp_dmc_data)
cebpd_exp_dmc_data$CEBPD <- as.numeric(cebpd_exp_dmc_data$CEBPD)

rvd <- cebpd_exp_dmc_data

#all samples
rvd_p <- ggplot(rvd, aes(DMC, CEBPD)) +
  geom_point() +
  theme_classic() +
  labs(x = "Mean of Differentially Methylated CpG (DMC) levels", y = "CEBPD Expression (log2FC)", title = "CEBPD Expression vs Mean DMC") +
  geom_smooth(method = lm, se = F, colour = "black") +
  #theme (plot_title = element_text(hjust = 0.5)) +
  #geom_text(x = 20, y = 10, label = paste("R = ", round(cor(rvd2$DMC, rvd2$CEBPD), 2))) +
  #theme(axis.line = element_line(color = "black", size = 0.5, axis.text.x = element_text(size = 30)))+
  theme( axis.text.x = element_text(size = 20), axis.text.y = element_text(size = 20), axis.title.x = element_text(size = 20), axis.title.y = element_text(size = 20)) +
  stat_cor(aes(DMC, CEBPD), method = "spearman", label.x = 0.3, label.y = 3, size = 7)
rvd_p
```

CEBPD expression vs specific DMC correlation

```{r}
rvdlog2fc <- read.table("DESeq2_Result_RvsD_logFC_and_qvalue.txt", header = T)
class(rvdlog2fc)
names(rvdlog2fc)
unique(rvdlog2fc$patientID)

rvdCEBPDlog2fc <- rvdlog2fc %>% filter (geneID == "CEBPD") %>% dplyr::select(patientID, RvsD_logFC)
col_names <- rvdCEBPDlog2fc$patientID
rvdCEBPDlog2fc <- t(rvdCEBPDlog2fc)
colnames(rvdCEBPDlog2fc) <- col_names
class(rvdCEBPDlog2fc)

rvdCEBPDlog2fc <- as.data.frame(rvdCEBPDlog2fc)
rvdCEBPDlog2fc <- rvdCEBPDlog2fc[2,]
row.names(rvdCEBPDlog2fc) <- "CEBPD"
rvdCEBPDlog2fc <- as.data.frame(t(rvdCEBPDlog2fc))


selectCpG_g <- g[row.names(g) %in% c("48650539", "48650533", "48650524"),]
selectCpG_g <- as.data.frame(t(selectCpG_g))
selectCpG_g_rownames <- row.names(selectCpG_g)
rvdCEBPDlog2fc <- subset(rvdCEBPDlog2fc, row.names(rvdCEBPDlog2fc) %in% selectCpG_g_rownames)

dim(selectCpG_g)
dim(rvdCEBPDlog2fc) 


merged_d <- merge(rvdCEBPDlog2fc, selectCpG_g, by = "row.names")
colnames(merged_d)

colnames(merged_d) <- c("patientID", "CEBPD_FC", "cpg.48650524",  "cpg.48650533",  "cpg.48650539")
merged_d[merged_d == 0] <- NA

ggplot(merged_d, aes(CEBPD_FC, cpg.48650539)) +
  geom_point() +
  theme_classic() +
  labs(x = "Mean of Differentially Methylated CpG (DMC) levels", y = "CEBPD Expression (log2FC)", title = "CEBPD Expression vs Mean DMC") +
  geom_smooth(method = lm, se = F, colour = "black") +
  #theme (plot_title = element_text(hjust = 0.5)) +
  #geom_text(x = 20, y = 10, label = paste("R = ", round(cor(rvd2$DMC, rvd2$CEBPD), 2))) +
  #theme(axis.line = element_line(color = "black", size = 0.5, axis.text.x = element_text(size = 30)))+
  theme( axis.text.x = element_text(size = 20), axis.text.y = element_text(size = 20), axis.title.x = element_text(size = 20), axis.title.y = element_text(size = 20)) +
  stat_cor(aes(CEBPD_FC, cpg.48650533), method = "spearman", label.x = 0.3, label.y = 3, size = 7, na.rm = TRUE)
```

Patients proportion with significantly methylated CpGs

```{r}
#rm(list = ls())

d <- read.csv("pAML_DMCs.csv")
d <- d[colSums(!is.na(d)) > 0]
allNA_CpG_removed <- d
row.names(allNA_CpG_removed) <- allNA_CpG_removed$start
allNA_CpG_removed <- allNA_CpG_removed %>% dplyr::select(-c(PAML59, X, start, SGUR3)) #SGUR3-CEBPD upregulation
patients_allNA_CpG_removed <- colnames(allNA_CpG_removed)

cebpdExp <- read.csv("pairedAML_RvsD_DiffExpGenes_logFC_1.2_qvalue_0.05_Filtered_2017-07-11.csv")
row.names(cebpdExp) <- cebpdExp$X
cebpdExp <- subset(cebpdExp, select = -c(X, TotalNum, UpRegulated, DownRegulated, Proportion.Of.Upregulated, Proportion.Of.Downegulated))
cebpdExp <- subset(cebpdExp, row.names(cebpdExp) == "CEBPD")
cebpdNoExp <- cebpdExp %>% select_if(is.na) 
patients_cebpdNoExp <- colnames(cebpdNoExp)

unchanged_cebpdExp_samples <- intersect(patients_cebpdNoExp, patients_allNA_CpG_removed)
dn_cebpdExp_samples <- setdiff(patients_allNA_CpG_removed, unchanged_cebpdExp_samples)
allNA_CpG_removed <- allNA_CpG_removed %>% dplyr::select(unchanged_cebpdExp_samples, dn_cebpdExp_samples)#col1:10 unchaged
df <- allNA_CpG_removed

prop_unch <- apply(df[,1:10], MARGIN = 1, function(x) sum(!is.na(x))/ncol(df))
prop_dn <- apply(df[,11:24], MARGIN = 1, function(x) sum(!is.na(x))/ncol(df))
df$prop_unch <- prop_unch
df$prop_dn <- prop_dn
df$prop_total <- prop_unch + prop_dn


df <- df[df$prop_total > 0,] #select for prop_total > 0.15
view(df)
dim(df)

df_plot <- data.frame(condition=rep(c("CEBPD_Unchanged", "CEBPD_Downregulated"), each=32),
                CpG_sites=rep(row.names(df), 2),
                prop=c(df$prop_unch, df$prop_dn))

p <- df_plot %>% ggplot(aes(x = CpG_sites, y = prop, fill = condition)) + geom_bar(stat = "identity", width = 0.5) + coord_flip() + theme_classic()
p + theme(axis.text.x = element_text(size = 15), axis.text.y = element_text(size = 15), axis.title.x = element_text(size = 15), axis.title.y = element_text(size = 15)) +
  labs(y = "Patients Proportion", x = "CpG Sites")

```

