---
title: "study_dataset_cebpd_mRNA_vs_750bp.promoter_methylation_correlation"
author: "SP"
date: "1/27/2023"
output: html_document
---

Load Packages

```{r}
library(tidyverse)
library(methylKit)
library(pheatmap)
library(RColorBrewer)
library(ggpubr)
library(MASS)
```

Codes to copy files from one directory to another directory

```{r}
to_dir <-"/Volumes/bioinfosp/paml_methylation/non_differential/test/"
from_dir <-'~/Desktop/r_practice/'
files_to_copy <- list.files(from_dir, pattern = '.csv')

for (i in files_to_copy){
  file.copy(paste0(from_dir, i), to_dir)
}

```

Renaming methylcall txt filenames in coherence with RNAseq ID names

```{r}
#cpg.PAML_UHU samples
lst <- list.files(pattern = "cpg.PAML_UHU")
for (i in lst) {
  print(paste("Renaming", i))
  file.rename(i, paste0(strsplit(i, "[._]")[[1]][2], strsplit(i, "[._]")[[1]][4], "_", strsplit(i, "[._]")[[1]][5], ".txt"))
}

#cpg.Sample_PAML_UHU samples
lst <- list.files(pattern = "cpg.Sample_PAML_UHU")
for (i in lst) {
  print(paste("Renaming", i))
  file.rename(i, paste0(strsplit(i, "[._]")[[1]][3], strsplit(i, "[._]")[[1]][5], "_", strsplit(i, "[._]")[[1]][6], ".txt"))
}

#methylCall_Samples
lst <- list.files(pattern = "methylCall_Sample")
for (i in lst) {
  print(paste("Renaming", i))
  file.rename(i, paste0(strsplit(i, "[._]")[[1]][3], strsplit(i, "[._]")[[1]][4], strsplit(i, "[._]")[[1]][5],"_", strsplit(i, "[._]")[[1]][6], ".txt"))
}

#cpg.PAML samples
lst <- list.files(pattern = "cpg.PAML")
for (i in lst) {
  print(paste("Renaming", i))
  file.rename(i, paste0(strsplit(i, "[._]")[[1]][2], strsplit(i, "[._]")[[1]][3], "_", strsplit(i, "[._]")[[1]][4], ".txt"))
}
```

Keep methylcall files matching with RNAseq sample IDs

```{r}
#Methylation file names
meth_ids <- list.files(path = '/Volumes/bioinfosp/paml_methylation/10X_methylcall_files-mod_names/', pattern = "*.txt")
'%!in%' <- Negate('%in%')
meth_ids <- meth_ids[meth_ids  %!in% c("AMLpairedDec2015_refseq_tpm.txt", "README.txt")]

#RNAseq sample ID list
tpm <- read.table('AMLpairedDec2015_refseq_tpm.txt', header = T)
row.names(tpm) <- tpm$id
tpm <- tpm[,colnames(tpm) != 'id']
tpm <- tpm[, !str_detect(colnames(tpm), pattern = 'NBM|replicate')]
tpm_ids <- names(tpm)

#mody sample ID names
tpm_ids_txt <- character()
for (i in 1:length(tpm_ids)){
  temp <- paste0(tpm_ids[i], '.txt')
  tpm_ids_txt[i] <- temp
}

tpm_ids <- tpm_ids_txt

#Common sample IDs between methylation and RNAseq data
com_ids <- intersect(meth_ids, tpm_ids) 

#Remove methylation data files for which there is no RNAseq data
remov_txt <- setdiff(meth_ids, com_ids)
unlink(remov_txt)
```

Extract methylattion data for CEBPD promoter coordinates from methylcall files

```{r}
filenames <- list.files(path = '/Volumes/bioinfosp/paml_methylation/10X_methylcall_files-mod_names/', pattern = "*.txt")
'%!in%' <- Negate('%in%')
filenames <- filenames[filenames[] %!in% c('AMLpairedDec2015_refseq_tpm.txt', 'README.txt')]
#filenames[filenames != 'README.txt']
outputnames <- str_remove(filenames, ".txt")
path <- '/Volumes/bioinfosp/paml_methylation/10X_methylcall_files-mod_names/'

for (i in 1:length(filenames)) {
  d <- read.table(file = paste0(path, filenames[i]), header = T)
  d <- d %>% filter(chr == "chr8") %>%
    filter(base > 48650520) %>% 
    filter(base < 48651270) 
  d <- d %>% rename_at("freqC", ~outputnames[i])
  write.csv(d, paste0(outputnames[i], ".csv"))
}
```

Merging all diagnosis patients CEBPD DMCs data

```{r}
pAML_files <- list.files(full.names = F, pattern = "_1.csv" )
output_files <- str_remove(pAML_files, ".csv")

c <- read.csv("PAML1_1.csv")
c <- c %>% 
  dplyr::select(contains("PAML") | contains("SGU"), "base")
for ( i in 2:length(pAML_files)){
  temp_data <- read.csv(pAML_files[i])
  temp_data <- temp_data %>% 
    dplyr::select(base, contains("PAML") | contains("SGU"))
  c <- merge(c, temp_data, by = "base", all.x = T, all.y = T)  #check all.x and all.y
  write.csv (c, paste0("pAML_diag_DMCs.csv"))
}
```

Merging all relapse patients CEBPD DMCs data

```{r}
pAML_files <- list.files(full.names = F, pattern = "_2.csv" )
output_files <- str_remove(pAML_files, ".csv")

c <- read.csv("PAML1_2.csv")
c <- c %>% 
  dplyr::select(contains("PAML") | contains("SGU"), "base")
for ( i in 2:length(pAML_files)){
  temp_data <- read.csv(pAML_files[i])
  temp_data <- temp_data %>% 
    dplyr::select(base, contains("PAML") | contains("SGU"))
  c <- merge(c, temp_data, by = "base", all.x = T, all.y = T)  #check all.x and all.y
  write.csv (c, paste0("pAML_rel_DMCs.csv"))
}
```

Common DMCs between diagnosis and relapse

```{r}
df_diagdmc <- read.csv('pAML_diag_DMCs.csv')
df_diagdmc <- drop_na(df_diagdmc)

df_reldmc <- read.csv('pAML_rel_DMCs.csv')
df_reldmc <- drop_na(df_reldmc)

com_dmc <- intersect(df_diagdmc$base, df_reldmc$base)

df_diagdmc <- df_diagdmc[df_diagdmc$base %in% com_dmc,]
df_relgdmc <- df_reldmc[df_reldmc$base %in% com_dmc,]

```

Mean DMC diagnosis patients

```{r}
row.names(df_diagdmc) <- df_diagdmc$base
"%!in%" <- Negate("%in%")
df_diagdmc <- df_diagdmc[, colnames(df_diagdmc) %!in% c("X", "base")]

df_diagdmc_paml <- df_diagdmc[, str_detect(colnames(df_diagdmc), pattern = "PAML")]
df_diagdmc_sg <- df_diagdmc[, str_detect(colnames(df_diagdmc), pattern = "SG")]

df_diagdmc_paml_mean <- as.data.frame(apply(df_diagdmc_paml, 2, mean))
colnames(df_diagdmc_paml_mean) <- 'meth_mean'

df_diagdmc_sg_mean <- as.data.frame(apply(df_diagdmc_sg, 2, mean))
colnames(df_diagdmc_sg_mean) <- 'meth_mean'
```

Mean DMC relapse patients

```{r}
row.names(df_reldmc) <- df_reldmc$base
df_reldmc <- df_reldmc[, colnames(df_reldmc) %!in% c("X", "base")]

df_reldmc_paml <- df_reldmc[, str_detect(colnames(df_reldmc), pattern = "PAML")]
df_reldmc_sg <- df_reldmc[, str_detect(colnames(df_reldmc), pattern = "SG")]

df_reldmc_paml_mean <- as.data.frame(apply(df_reldmc_paml, 2, mean))
colnames(df_reldmc_paml_mean) <- 'meth_mean'

df_reldmc_sg_mean <- as.data.frame(apply(df_reldmc_sg, 2, mean))
colnames(df_reldmc_sg_mean) <- 'meth_mean'
```

Extract CEBPD tpms from RNAseq data for which methylation data available

```{r}
meth_ids <- list.files(path = '/Volumes/bioinfosp/paml_methylation/10X_methylcall_files-mod_names/', pattern = "*.txt")
'%!in%' <- Negate('%in%')
meth_ids <- meth_ids[meth_ids  %!in% c("AMLpairedDec2015_refseq_tpm.txt", "README.txt")]
length(meth_ids)
tpm <- read.table('AMLpairedDec2015_refseq_tpm.txt', header = T)
row.names(tpm) <- tpm$id
tpm <- tpm[,colnames(tpm) != 'id']
tpm <- tpm[, !str_detect(colnames(tpm), pattern = 'NBM|replicate')]
tpm_ids <- names(tpm)
tpm_ids_txt <- character()
for (i in 1:length(tpm_ids)){
  temp <- paste0(tpm_ids[i], '.txt')
  tpm_ids_txt[i] <- temp
}
tpm_ids <- tpm_ids_txt
com_ids <- intersect(meth_ids, tpm_ids) 



tpm_cebpd <- tpm[row.names(tpm) == 'CEBPD',]

temp_ids <- character()
for (i in 1:length(com_ids)){
  temp <- str_remove(com_ids[i], '.txt')
  temp_ids[i] <- temp
}
com_ids <- temp_ids

tpm_cebpd <- tpm_cebpd[, colnames(tpm_cebpd) %in% com_ids]
tpm_cebpd <- as.data.frame(t(tpm_cebpd))
colnames(tpm_cebpd) <- 'tpm'

tpm_diag <- subset(tpm_cebpd, str_detect(row.names(tpm_cebpd), pattern = "_1"))
tpm_diag_paml <- subset(tpm_diag, str_detect(row.names(tpm_diag), pattern = "PAML"))
tpm_diag_sg <- subset(tpm_diag, str_detect(row.names(tpm_diag), pattern = "SG"))


tpm_rel <- subset(tpm_cebpd, str_detect(row.names(tpm_cebpd), pattern = "_2"))
tpm_rel_paml <- subset(tpm_rel, str_detect(row.names(tpm_rel), pattern = "PAML"))
tpm_rel_sg <- subset(tpm_rel, str_detect(row.names(tpm_rel), pattern = "SG"))

```

Get IDs of the 59 study patients

```{r}
ft9  <- read.csv('pairedAML_RvsD_DiffExpGenes_logFC_1.2_qvalue_0.05_Filtered_2017-07-11_SP1.csv')
row.names(ft9) <- ft9$X
ft9 <- ft9[, colnames(ft9) != 'X']
ft9 <- ft9[,1:59]

ft9_diag_names <- character()
for (i in 1:length(colnames(ft9))){
  temp <- paste0(colnames(ft9)[i], '_1')
  ft9_diag_names[i] <- temp
}
ft9_diag_names

ft9_rel_names <- character()
for (i in 1:length(colnames(ft9))){
  temp <- paste0(colnames(ft9)[i], '_2')
  ft9_rel_names[i] <- temp
}
ft9_rel_names

#diagnosis samples to be removed
d <- c(row.names(df_diagdmc_paml_mean), row.names(df_diagdmc_sg_mean))
d_remove <- setdiff(d, ft9_diag_names)
d_remove
#relapse samples to be removed
r <- c(row.names(df_reldmc_paml_mean), row.names(df_reldmc_sg_mean))
r_remove <- setdiff(r, ft9_rel_names)
r_remove

```

tpm vs methylation mean correlation (n = 59)

```{r}
#Samples to be excluded
#DIAGNOSIS
diag_paml_merged <- merge(tpm_diag_paml, df_diagdmc_paml_mean, by = 'row.names')
row.names(diag_paml_merged) <- diag_paml_merged$Row.names
diag_paml_merged <- diag_paml_merged[,colnames(diag_paml_merged) != 'Row.names']
diag_paml_merged <- diag_paml_merged[(row.names(diag_paml_merged) != 'PAML62_1'),]

ggplot(diag_paml_merged, aes(log(tpm), log(meth_mean))) +
  geom_point() +
  theme_classic() +
  labs(x = "log (CEBPD (TPM))", y = "log (Meth_mean)") +
  geom_smooth(method = lm, se = F, colour = "black") +
  #theme (plot_title = element_text(hjust = 0.5)) +
  #geom_text(x = 20, y = 10, label = paste("R = ", round(cor(rvd2$DMC, rvd2$CEBPD), 2))) +
  #theme(axis.line = element_line(color = "black", size = 0.5, axis.text.x = element_text(size = 30)))+
  theme( axis.text.x = element_text(size = 34), axis.text.y = element_text(size = 34), axis.title.x = element_text(size = 34), axis.title.y = element_text(size = 34)) +
  stat_cor(aes(tpm, meth_mean), method = "spearman",  label.y = 4, size = 10.8, label.x = 2)


diag_sg_merged <- merge(tpm_diag_sg, df_diagdmc_sg_mean, by = 'row.names')
row.names(diag_sg_merged) <- diag_sg_merged$Row.names
diag_sg_merged <- diag_sg_merged[,colnames(diag_sg_merged) != 'Row.names']
diag_sg_merged <- diag_sg_merged[(row.names(diag_sg_merged) %!in% c("SGUA11_1", "SGUA14_1", "SGUP21_1", "SGUR4_1")),]

ggplot(diag_sg_merged, aes(log(tpm), log(meth_mean))) +
  geom_point() +
  theme_classic() +
  labs(x = "log (CEBPD (TPM))", y = "log (Meth_mean)") +
  geom_smooth(method = lm, se = F, colour = "black") +
  #theme (plot_title = element_text(hjust = 0.5)) +
  #geom_text(x = 20, y = 10, label = paste("R = ", round(cor(rvd2$DMC, rvd2$CEBPD), 2))) +
  #theme(axis.line = element_line(color = "black", size = 0.5, axis.text.x = element_text(size = 30)))+
  theme( axis.text.x = element_text(size = 34), axis.text.y = element_text(size = 34), axis.title.x = element_text(size = 34), axis.title.y = element_text(size = 34)) +
  stat_cor(aes(tpm, meth_mean), method = "spearman",  label.y = 4, size = 10.8, label.x = 0.8)




#RELAPSE
rel_paml_merged <- merge(tpm_rel_paml, df_reldmc_paml_mean, by = 'row.names')
row.names(rel_paml_merged) <- rel_paml_merged$Row.names
rel_paml_merged <- rel_paml_merged[,colnames(rel_paml_merged) != 'Row.names']

ggplot(rel_paml_merged, aes(log(tpm), log(meth_mean))) +
  geom_point() +
  theme_classic() +
  labs(x = "log (CEBPD (TPM))", y = "log (Meth_mean)") +
  geom_smooth(method = lm, se = F, colour = "black") +
  #theme (plot_title = element_text(hjust = 0.5)) +
  #geom_text(x = 20, y = 10, label = paste("R = ", round(cor(rvd2$DMC, rvd2$CEBPD), 2))) +
  #theme(axis.line = element_line(color = "black", size = 0.5, axis.text.x = element_text(size = 30)))+
  theme( axis.text.x = element_text(size = 34), axis.text.y = element_text(size = 34), axis.title.x = element_text(size = 34), axis.title.y = element_text(size = 34)) +
  stat_cor(aes(tpm, meth_mean), method = "spearman",  label.y = 4, size = 10.8, label.x = 2)


rel_sg_merged <- merge(tpm_rel_sg, df_reldmc_sg_mean, by = 'row.names')
row.names(rel_sg_merged) <- rel_sg_merged$Row.names
rel_sg_merged <- rel_sg_merged[,colnames(rel_sg_merged) != 'Row.names']
rel_sg_merged <- rel_sg_merged[(row.names(rel_sg_merged) %!in% c("SGUA9_2", "SGUP14_2", "SGUP30_2")),]


ggplot(rel_sg_merged, aes(log(tpm), log(meth_mean))) +
  geom_point() +
  theme_classic() +
  labs(x = "log (CEBPD (TPM))", y = "log (Meth_mean)") +
  geom_smooth(method = lm, se = F, colour = "black") +
  #theme (plot_title = element_text(hjust = 0.5)) +
  #geom_text(x = 20, y = 10, label = paste("R = ", round(cor(rvd2$DMC, rvd2$CEBPD), 2))) +
  #theme(axis.line = element_line(color = "black", size = 0.5, axis.text.x = element_text(size = 30)))+
  theme( axis.text.x = element_text(size = 34), axis.text.y = element_text(size = 34), axis.title.x = element_text(size = 34), axis.title.y = element_text(size = 34)) +
  stat_cor(aes(tpm, meth_mean), method = "spearman",  label.y = 4, size = 10.8, label.x = 0.8)
```